---
title: "tmtUvDose_Extras"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r 02_Mapping-functions}
# ------------------------------------------------------------------------------------------------------------
# Step 02: Mapping 'background' as well as 'expressed' list of proteins to various annotations
# For full list of query fields, see http://mygene.info/clients/ and 'Available fields'

# ------------------------------------------------------------------------------------------------------------
# Function  : 'myProtMapper' to use the function 'queryMany' from Bioconductor package mygene as fast and most up-to-date
# Input 
#       : ids = a character list of ids which can be uniprot, ensembl gene, gene symbol,etc
#       : id.type = what type of ids have you provided in the 'ids' list. Default = "uniprot"
#       : outlist = list of ids you want as an output. Default = c("interpro","ensembl.gene","go")
#       : modify = Logical, Default = T; Would you like to modify fields such as interpro, enseml, go to make them more human readable.
# Output  : A dataframe with required ids and input ids 
# ------------------------------------------------------------------------------------------------------------------------

myProtMapper <- function(ids,id.type="uniprot",out.fields=c("interpro.short_desc","ensembl.gene","go.MF.id","go.CC.id","go.BP.id"),modify=T){
  
  # Get the mapping
  qm = queryMany(ids,scopes=id.type,fields=out.fields)
  
  # Returning variable
  ret.qm = NULL
  
  # Resolve the mappings to make them human readable
  if(modify == T){
    qm$go.all = NULL
    
    # Interpro mappings
    if(is.element("interpro",colnames(qm))){
      qm$domains = sapply(qm$interpro,function(x) paste(unlist(x),collapse=";"))
    }
    else{
      print("No Interpro domains")
    }
    
    # GO mappings
    if(!is.na(grep("go",colnames(qm)))){
      
      # Grep all the go columns 'go.CC','go.MF','go.BP'
      f = grep("go",colnames(qm), value=T)
      
      for(i in f){
        print(i)
        temp = sapply(qm[,i],function(x) paste(unlist(x),collapse=";"))
        if(temp!=""){
          if(is.null(qm$go.all)){
            qm$go.all = temp  
          }
          else{
             qm$go.all = paste(qm$go.all,temp,sep=";")
          }
        }else{
          print(paste("missing",i),sep=" ")
        }
      }
      # Replace any extra separation marks
      #qm$go.all = gsub("^;","",gsub(";;","",qm$go.all))
      qm$go.all = gsub("^;","",gsub(";;",";",qm$go.all))
    }
    else{
      print("No GO terms")
    }
    
    # Ensembl.gene mappings
    if(is.element("ensembl",colnames(qm))){
      qm$ens = sapply(qm$ensembl,function(x) paste(unlist(x),collapse=";"))
    }
    else{
      print("No Ensembl Ids")
    }
    
    # Return mapped structure with tidy columns
    ret.qm = qm
  }
  else{
    ret.qm = qm
  }
  
  return(data.frame(ret.qm))
}


# ------------------------------------------------------------------------------------------------------------
# Function  : 'makeGene2Cat' to produce a 1:1 mapping of uniprot/ensembl/symbols to GO/Interpro terms. 
#              Will be used as input into the 'goseq' function in the gene2cat slot
# Input 
#           : dat = dataframe with ids and go/interpro terms (obtained from myProtMapper)
#           : from.id =  ids you want to map 'from'. Default = "uniprot"
#           : to.id =  ids you want to map to c("interpro","ensembl.gene","go")
#           : splt = symbol you want to split by if there are multiple ids
# Output  : A two column dataframe with Uniprot ids in the first and Go/Interpro in the second
# ------------------------------------------------------------------------------------------------------------------------

makeGene2Cat <- function(dat,from.id,to.id,splt){
  
  cat.frame = dat[,c(from.id,to.id)]
  d.dt = data.table(cat.frame,key=colnames(cat.frame[,from.id]))
  cat.out = data.frame(d.dt[, list(to.id = unlist(strsplit(get(to.id), splt))), by=from.id])
  cat.out = unique(cat.out)
  
  return(cat.out)
}

# ------------------------------------------------------------------------
# Function  : runGoseq
# Aim       : runs goseq on a list of genes
# Input     : list of genes
# Output    : enriched list of Interpro domains
# ------------------------------------------------------------------------

runGoseq <- function(genelist,bglist,bias=NULL,cat.oligo=NULL,signif=0.05){

  # setting up goseq object
  all.genes.comp = rep(0,nrow(bglist))
  names(all.genes.comp) = rownames(bglist)
  all.genes.comp[which(names(all.genes.comp) %in% unique(genelist))] = 1
  table(all.genes.comp)
  
  # Remove missing values
  comp.no.missing = all.genes.comp[which(!is.na(names(all.genes.comp)))]
  table(comp.no.missing)
    
  # Running the function to calculate weights - with user provided bias or default gene length bias
  pwf.comp = nullp(comp.no.missing,bias.data = bias)
  
  # Running goseq enrichment using default mapping within goseq or user provided mapper
  goseq.comp.cat = goseq(pwf.comp,gene2cat = cat.oligo)
  goseq.comp.cat$BH_over_represented_pvalue = p.adjust(goseq.comp.cat$over_represented_pvalue,method = "BH")
  enriched.goseq.comp.cat = goseq.comp.cat[which(goseq.comp.cat$BH_over_represented_pvalue <= signif),]
  
  return(list(goseq.comp.cat,enriched.goseq.comp.cat))

}
```
```{r}
#both= intersect(Trizol, oligodT)
both.enrich.go = runGoseq(both,bias.df, bias.df$protbias,geiger.cat.go)
both.enrich.go[[2]][,c(1,6:7,4:5,2,8)]
both.enrich.interpro = runGoseq(both,bias.df, bias.df$protbias,geiger.cat)
#write.table(both.enrich,paste(outdir,"oligodT-Trizol-common-genes-Interpro-enrichment.txt",sep="/"),sep="\t",row.names=F,quote=F)

#oligo.only = attr(v,"intersection")$`oligodT` # n = 57
oligo.only = setdiff(oligodT,Trizol) # n = 57
oligo.only.enrich.go = runGoseq(oligo.only,bias.df, bias.df$protbias,geiger.cat.go)
oligo.only.enrich.go[,c(1,6:7,4:5,2,8)]
oligo.only.enrich.interpro = runGoseq(oligo.only,bias.df, bias.df$protbias,geiger.cat)
#write.table(oligo.only.enrich,paste(outdir,"oligodT-only-genes-101-Interpro-enrichment.txt",sep="/"),sep="\t",row.names=F,quote=F)

#trizol.only = attr(v,"intersection")$`Trizol` # n = 1543
trizol.only = setdiff(Trizol,oligodT)
trizol.only.enrich.go = runGoseq(trizol.only,bias.df, bias.df$protbias,geiger.cat.go)
trizol.only.enrich.go[,c(1,6:7,4:5,2,8)]
trizol.only.enrich.interpro = runGoseq(trizol.only,bias.df, bias.df$protbias,geiger.cat)
#write.table(trizol.only.enrich,paste(outdir,"Trizol-only-genes-1296-Interpro-enrichment.txt",sep="/"),sep="\t",row.names=F,quote=F)
```

```{r}
# Convert data table
# geiger.doms.sym = data.frame(geiger.qm[,c("symbol","domains")])
# geiger.gos.sym = data.frame(geiger.qm[,c("symbol","go.all")])

# d.dt.sym <- data.table(geiger.doms.sym, key="symbol")
# geiger.cat.sym <- d.dt.sym[, list(domains = unlist(strsplit(domains, "; "))), by=symbol]

# go.dt.sym = data.table(geiger.gos.sym, key="symbol")
# geiger.cat.go.sym <- go.dt.sym[, list(go.terms = unlist(strsplit(go.all, "; "))), by=symbol]

# Setting up a 'genes' vector. We've only kept those genes present in u2os as the universe
# Then we mark all those that are enriched in the analysis as genes of interest (DE if you will)
# We have 1435 genes out of 1516 that have bias data and go terms

# Running goseq for oligodT which have gene symbols as identifiers
# rownames(bias.df) = bias.df$SYMBOL
# oligo crosslinked
cl.qm = queryMany(oligo.cl,scopes="uniprot",fields=c("ensembl","name","symbol","interpro","go"),species=9606)
cl.qm$domains = sapply(sapply(cl.qm$interpro,"[[",3),function(x) paste(x,collapse="; "))
cl.qm$num.rbds = rowSums(sapply(rbd.doms, function(x) grepl(x, cl.qm$domains)))
cl.qm$which.rbd = apply(sapply(rbd.doms, function(x) grepl(x, cl.qm$domains)),1, function(y) paste(names(which(y==T)),collapse="; "))
table(cl.qm$num.rbds)
paste("Percentage of proteins with RNA-binding domains in cl = ",round(100*sum(table(cl.in.nc.qm$num.rbds)[2:4])/sum(table(cl.in.nc.qm$num.rbds)),2),sep="")

# oligo non-crosslinked
nc.qm = queryMany(oligo.nc$GeneName,scopes="symbol",fields=c("ensembl","name","uniprot","interpro","go"),species=9606)
nc.qm$domains = sapply(sapply(nc.qm$interpro,"[[",3),function(x) paste(x,collapse="; "))
nc.qm$num.rbds = rowSums(sapply(rbd.doms, function(x) grepl(x, nc.qm$domains)))
nc.qm$which.rbd = apply(sapply(rbd.doms, function(x) grepl(x, nc.qm$domains)),1, function(y) paste(names(which(y==T)),collapse="; "))

# oligo cl not in nc
cl.only.qm = queryMany(oligo.cl$GeneName,scopes="symbol",fields=c("ensembl","name","uniprot","interpro","go"),species=9606)
cl.only.qm$domains = sapply(sapply(cl.only.qm$interpro,"[[",3),function(x) paste(x,collapse="; "))
cl.only.qm$num.rbds = rowSums(sapply(rbd.doms, function(x) grepl(x, cl.only.qm$domains)))
cl.only.qm$which.rbd = apply(sapply(rbd.doms, function(x) grepl(x, cl.only.qm$domains)),1, function(y) paste(names(which(y==T)),collapse="; "))

# Number of proteins with known RBDs
table(uv.qm$num.rbds) # 315/1780 (18%) that have one or more known RBD domains
table(geiger.qm$num.rbds) # 707/7574 (9%) that have one or more known RBD domains
table(cl.qm$num.rbds) # 127/332 (38%) that have one or more known RBD domains
table(nc.qm$num.rbds) # 28/74 (38%) that have one or more known RBD domains
table(cl.only.qm$num.rbds) # 99/261 (38%) that have one or more known RBD domains

# Looking at duplicated Ensembl ids
dups = bias.df.ens.sp[duplicated(bias.df.ens.sp$ens),] 

# Data without duplicates
bias.no.dups = bias.df.ens.sp[!duplicated(bias.df.ens.sp$ensembl),]
rownames(bias.no.dups) = bias.no.dups$ensembl
```


```{r 08_Filtering-oligodT-proteins}

# -----------------------------------
# Some filtering of oligodT proteins
#------------------------------------

# Filtering for those with Swissprot ids only
#--------------------------------------------

# 587/762 cross-linked proteins are high confidence and are not missing SwissProt ids
oligo.cl.sp = oligo.cl.qm[which(!is.na(oligo.cl.qm$uniprot.Swiss.Prot)),]
length(unique(oligo.cl.sp$uniprot.Swiss.Prot))

# 176/224 non-crosslined proteins are high confidence
oligo.nc.sp = oligo.nc.qm[which(!is.na(oligo.nc.qm$uniprot.Swiss.Prot)),]
length(unique(oligo.nc.sp$uniprot.Swiss.Prot))

# Remove mass spec contaminant proteins as we did with Trizol data
#--------------------------------------------------------------------
oligo.cl.no.contam = oligo.cl.sp[-which(oligo.cl.sp$uniprot.Swiss.Prot %in% contam$Protein.Group.Accessions),]
length(unique(oligo.cl.no.contam$uniprot.Swiss.Prot)) # 586 so we removed 1 contaminants ALB not present in 'nc' samples

oligo.nc.no.contam = oligo.nc.sp[-which(oligo.nc.sp$uniprot.Swiss.Prot %in% contam$Protein.Group.Accessions),]
length(unique(oligo.nc.no.contam$uniprot.Swiss.Prot)) # 172 so we removed 4 contaminants KRT2, KRT10, RPS27A, KRT1

```
We have two datasets now - oligo.cl.no.contam made of 586 unique proteins and oligo.nc.no.contam made of 172 proteins. Next step is to set up goseq. In the initial run, I forgot that the bias data was from U2OS Geiger rather than our own expression so will implement this.

In addition, we annotate proteins given by their GeneNames/HGNC Symbols to entrezgene id, uniprot id, gene description, interpro domains and trembl ids. Because one gene can be alternatively spliced to multiple proteins, we have a one-to-many mapping and 258 crosslinked genes give us 762 crosslinked proteins. Similarly, 73 non-crosslinked genes give us 224 non-crosslinked proteins 

```{r 09_oligodT-GO-Interpro-enrichment}

# ------------------------------------------------------------------------
# Function  : runGoseq
# Aim       : runs goseq on a list of genes
# Input     : list of genes
# Output    : enriched list of Interpro domains
# ------------------------------------------------------------------------

runGoseq <- function(genelist,bglist,bias.dat=NULL,cat.oligo){

  # setting up goseq object
  all.genes.comp = rep(0,nrow(bglist))
  names(all.genes.comp) = rownames(bglist)
  all.genes.comp[which(names(all.genes.comp) %in% unique(genelist))] = 1
  table(all.genes.comp)
  
  # Remove missing values
  comp.no.missing = all.genes.comp[which(!is.na(names(all.genes.comp)))]
  table(comp.no.missing)
  
    
  # Running the function to calculate weights. We have no bias information as we did in UV experiment
  # This is because mass spec was run in detection mode not quantitation mode. 
  pwf.comp = nullp(comp.no.missing,'hg19','geneSymbol', bias=bias.dat,plot.fit=TRUE)
  
  # goseq enrichment with domains for cross-linked samples
  goseq.comp = goseq(pwf.comp,gene2cat = cat.oligo)
  goseq.comp$BH_over_represented_pvalue = p.adjust(goseq.comp$over_represented_pvalue,method = "BH")
  goseq.comp
  
  enriched.goseq.comp = goseq.comp[which(goseq.comp$BH_over_represented_pvalue <= 0.05),]
  return(enriched.goseq.comp)

}

# -----------------------------------------------------------------
# Running goseq for Interpro domain enrichment (using UniProt IDs)
# -----------------------------------------------------------------

dt.oligo <- data.table(as.data.frame(geiger.qm[,c("query","domains")]), key="query")
cat.oligo <- dt.oligo[, list(domains = unlist(strsplit(domains, "; "))), by=query]

enriched.dom.cl.oligo = runGoseq(oligo.cl.no.contam$uniprot.Swiss.Prot, bias.df, bias.df$protbias, cat.oligo)
enriched.dom.nc.oligo = runGoseq(oligo.nc.no.contam$uniprot.Swiss.Prot, bias.df, bias.df$protbias, cat.oligo)


# --------------------------------------------------------
# Running goseq for GO enrichment (using Gene Symbols)
# -------------------------------------------------------

# Need to get gene symbols for u2os list
protbias.sym = rowMax(exprs(agg.u2os))
names(protbias.sym) = sapply(strsplit(as.character(fData(agg.u2os)$protein_description),"\\||\\_"),"[[",3)
length(protbias.sym)

bias.df.sym = data.frame(protbias.sym)
bias.df.sym$SYMBOL = rownames(bias.df.sym)
head(bias.df.sym)

cat.sym <- bg.list.bm[,c("hgnc_symbol","go_id")]

# Call runGoseq function
enriched.go.cl.oligo = runGoseq(oligo.cl.no.contam$query, bias.df.sym, bias.df.sym$protbias.sym, cat.sym)
enriched.go.cl.oligo = enriched.go.cl.oligo[,c(1,6,7,4:5,2,8,3)]
enriched.go.nc.oligo = runGoseq(oligo.nc.no.contam$query, bias.df.sym, bias.df.sym$protbias.sym, cat.sym)
enriched.go.nc.oligo = enriched.go.nc.oligo[,c(1,6,7,4:5,2,8,3)]

```

```{r 11_Extras}
#-------------------
# goseq with KEGG
#-------------------

# Restricted to only those proteins that had KEGG mappings in bioMart so not very extensive.
# Better mapping and analysis with clusterProfiler....

#------------------------------
# Function : mapPathwayToName
#------------------------------

mapPathwayToName <- function(organism) {
  KEGG_PATHWAY_LIST_BASE <- "http://rest.kegg.jp/list/pathway/"
  pathway_list_REST_url <- paste(KEGG_PATHWAY_LIST_BASE, organism, sep="")
 
  pathway_id_name <- data.frame()
 
  for (line in readLines(pathway_list_REST_url)) {
    tmp <- strsplit(line, "\t")[[1]]
    pathway_id <- strsplit(tmp[1], organism)[[1]][2]
    pathway_name <- tmp[2]
    pathway_name <- strsplit(pathway_name, "\\s+-\\s+")[[1]][1]
    pathway_id_name[pathway_id, 1] = pathway_name
 
  }
 
  names(pathway_id_name) <- "pathway_name"
  pathway_id_name
}

kegg.names = mapPathwayToName('hsa')
kegg.list = bg.list.bm[which(bg.list.bm$kegg_enzyme != ""),]
kegg.list$kegg = sapply(strsplit(kegg.list$kegg_enzyme,"\\+"),"[[",1)

# Check for GO enrichment given the bias data using KEGG mappings
kegg.all.genes = all.genes[which(names(all.genes) %in% kegg.list$uniprotswissprot)]
table(kegg.all.genes)
kegg.bias.df = bias.df[which(bias.df$UNIPROT %in% kegg.list$uniprotswissprot),]

# Go seq based on KEGG 
#write.table(data.frame(all.genes),"All-genes-for-goseq.txt",sep="\t",row.names=T, quote=F)
kegg.ids = read.table("Input/All-genes-for-goseq-with-KEGG.txt",sep="\t",header=T)

KEGG.pwf = nullp(all.genes,'hg19','ensGene', bias.data=as.numeric(bias.df$protbias),plot.fit=TRUE)
KEGG.wall = goseq(KEGG.pwf,gene2cat = kegg.ids[,c(1,2)])

KEGG.wall$BH_over_represented_pvalue = p.adjust(KEGG.wall$over_represented_pvalue,method = "BH")
KEGG.wall$description = kegg.names[KEGG.wall$category,]
KEGG.wall = KEGG.wall[,c(1,7,4:5,2,6,3)]
KEGG.wall

prot.list = as.character(unique(prot.list.bm$entrezgene)) # n = 1516
bg.list = as.character(unique(bg.list.bm$entrezgene)) # n = 7477

ggo.bp <- enrichGO(gene = prot.list,universe = bg.list,OrgDb = org.Hs.eg.db,ont = "BP",qvalueCutoff = 0.05,readable = TRUE)
ggo.mf <- enrichGO(gene = prot.list,universe = bg.list,OrgDb = org.Hs.eg.db,ont = "MF",qvalueCutoff = 0.05,readable = TRUE)
ggo.cc <- enrichGO(gene = prot.list,universe = bg.list,OrgDb = org.Hs.eg.db,ont = "CC",qvalueCutoff = 0.05,readable = TRUE)

ggo = rbind(data.frame(ggo.bp,Ontology="BP"),data.frame(ggo.mf,Ontology="MF"),data.frame(ggo.cc,Ontology="CC"))
ggo = ggo[,c(1:2,10,3:4,7,5:6)]
ggo
dotplot(ggo.mf)
dotplot(ggo.bp)
dotplot(ggo.cc)
head(ggo,n=20)


# Doing a KEGG based enrichment for the enriched protein list
kegg.prot = enrichKEGG(gene = prot.list,universe = bg.list,qvalueCutoff = 0.05)
kegg.prot = data.frame(kegg.prot)
kegg.prot = kegg.prot[order(kegg.prot$Count,decreasing=T),]

#-------------------------------------------------------------------------
# Mapping Geiger U2OS list of proteins to gene domains from Interpro.
#-------------------------------------------------------------------------

# MyGene.info can be accessed through python as well. 
library(mygene)

# Example of obtaining gene domains for each gene
#qm = queryMany(c("A0AVT1","A1L020","A1X283","A5YKK6","A6NHR9","A8MYA2"),scopes="uniprot",fields=c("id","entrezgene","name","symbol","interpro"))
#qm$domains = sapply(sapply(qm$interpro,"[[",3),function(x) paste(x,collapse="; "))
#head(qm)               

# Geiger list being annotated with interpro descriptions - fast as done in chunks
# 7414 uniprot IDs yield 7480 lines of data
geiger.qm = queryMany(unique(bg.list.bm$uniprotswissprot),scopes="uniprot",fields=c("ensembl","entrezgene","name","symbol","interpro"))
geiger.qm$domains = sapply(sapply(geiger.qm$interpro,"[[",3),function(x) paste(x,collapse="; "))

#-------------------------------------------------------------------------
# Mapping Trizol-enriched list of proteins to gene domains from Interpro.
#-------------------------------------------------------------------------
# 1495 uniprot IDs yield 1521 lines of data
uv.qm = queryMany(unique(prot.list.bm$uniprotswissprot),scopes="uniprot",fields=c("ensembl","entrezgene","name","symbol","interpro"))
uv.qm$domains = sapply(sapply(uv.qm$interpro,"[[",3),function(x) paste(x,collapse="; "))

# 7507 proteins map to 183612 annotated lines
#bg.list.bm = getBM(attributes = c("uniprotswissprot","ensembl_gene_id","hgnc_symbol","go_id","name_1006","interpro","interpro_short_description","interpro_description"),filters = "uniprotswissprot",values = fData(agg.u2os)$master_protein,mart = ensembl)
saveRDS(bg.list.bm, file="u2os.bg.list.Rds")

bg.list.bm = readRDS("u2os.bg.list.Rds")
head(bg.list.bm)
dim(bg.list.bm)
length(unique(bg.list.bm$hgnc_symbol)) # 7443

#------------------------------------------------------
# Mapping Trizol-enriched list of proteins to GO terms
#------------------------------------------------------

# 7507 proteins map to 183612 annotated lines
prot.list.bm = getBM(attributes = c("uniprotswissprot","ensembl_gene_id","hgnc_symbol","go_id","name_1006","interpro","interpro_short_description","interpro_description"),filters = "uniprotswissprot",values = fData(qnt.prot)$Master.Protein.Accessions,mart = ensembl)
saveRDS(bg.list.bm, file="uv.list.Rds")

prot.list.bm = readRDS("uv.list.Rds")
head(prot.list.bm)
dim(prot.list.bm) # 46005
length(unique(prot.list.bm$hgnc_symbol)) # 1518

# Need to map Geiger et al to GO terms, domains
library(biomaRt)
ensembl = useMart("ensembl", dataset = "hsapiens_gene_ensembl")

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
